---
title: "Predicting Rushing and Receiving Yards for NFL Players"
author: "paul trusela"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
#install.packages(c("nflfastR", "dplyr", "tidyr", "caret"))
#install.packages("gsisdecoder")
# install.packages("tidyr")
# install.packages("cluster")
# install.packages("factoextra")
# install.packages("sparcl")
# install.packages("teamcolors")
# install.packages("ggimage")
# install.packages("ggdark")
# install.packages("ggplot2")
# install.packages("gridExtra")
# install.packages("gt")
library(gt)
library(ggplot2)
library(gridExtra)
library(tidyr) 
library(cluster) 
library(factoextra) 
library(sparcl) 
library(teamcolors) 
library(ggimage)
library(nflfastR)
library(dplyr)
library(caret)
library(gsisdecoder)
library(tidyverse)
library(randomForest)


```

```{r}

pbp_data <- load_pbp(2023:2024)

```

CLEANING DATA
```{r}
player_data <- pbp_data %>%
  filter(play_type %in% c("run", "pass")) %>%
  mutate(
    player_name = case_when(
      play_type == "run" ~ rusher_player_name,
      play_type == "pass" ~ receiver_player_name,
      TRUE ~ NA_character_
    ),
    yards_gained = case_when(
      play_type == "run" ~ rushing_yards,
      play_type == "pass" ~ receiving_yards,
      TRUE ~ NA_real_
    ),
    opponent = defteam
  ) %>%
  filter(!is.na(player_name))

player_averages <- player_data %>%
  group_by(player_name) %>%
  summarise(
    avg_rushing_yards_per_game = sum(rushing_yards, na.rm = TRUE) / n_distinct(game_id),
    avg_receiving_yards_per_game = sum(receiving_yards, na.rm = TRUE) / n_distinct(game_id)
  )

player_data <- player_data %>%
  left_join(player_averages, by = "player_name") %>%
  group_by(player_name, week, opponent) %>%
  summarise(
    total_rushing_yards = sum(rushing_yards, na.rm = TRUE),
    total_receiving_yards = sum(receiving_yards, na.rm = TRUE),
    avg_rushing_yards_per_game = first(avg_rushing_yards_per_game),
    avg_receiving_yards_per_game = first(avg_receiving_yards_per_game)
  ) %>%
  ungroup()

player_data <- na.omit(player_data)

```


```{r}
set.seed(42)
train_index_rush <- createDataPartition(player_data$total_rushing_yards, p = 0.8, list = FALSE)
train_data_rush <- player_data[train_index_rush, ]
test_data_rush <- player_data[-train_index_rush, ]

set.seed(42)
train_index_rec <- createDataPartition(player_data$total_receiving_yards, p = 0.8, list = FALSE)
train_data_rec <- player_data[train_index_rec, ]
test_data_rec <- player_data[-train_index_rec, ]
```

RANDOM FORREST MODEL
```{r}

rf_model_rush <- train(
  total_rushing_yards ~ avg_rushing_yards_per_game + opponent,
  data = train_data_rush,
  method = "rf",
  trControl = trainControl(method = "cv", number = 5)
)

rf_model_rec <- train(
  total_receiving_yards ~ avg_receiving_yards_per_game + opponent,
  data = train_data_rec,
  method = "rf",
  trControl = trainControl(method = "cv", number = 5)
)


```

```{r}

rf_predictions_rush <- predict(rf_model_rush, test_data_rush)

rf_mae_rush <- mean(abs(rf_predictions_rush - test_data_rush$total_rushing_yards))
print(paste("Mean Absolute Error for Rushing Yards:", rf_mae_rush))

rf_predictions_rec <- predict(rf_model_rec, test_data_rec)

rf_mae_rec <- mean(abs(rf_predictions_rec - test_data_rec$total_receiving_yards))
print(paste("Mean Absolute Error for Receiving Yards:", rf_mae_rec))

```

```{r}
library(caret)

rushing_threshold <- 50
receiving_threshold <- 50

rf_binary_predictions_rush <- ifelse(rf_predictions_rush > rushing_threshold, 1, 0)
rf_binary_actual_rush <- ifelse(test_data_rush$total_rushing_yards > rushing_threshold, 1, 0)

rf_binary_predictions_rec <- ifelse(rf_predictions_rec > receiving_threshold, 1, 0)
rf_binary_actual_rec <- ifelse(test_data_rec$total_receiving_yards > receiving_threshold, 1, 0)

rf_conf_matrix_rush <- confusionMatrix(as.factor(rf_binary_predictions_rush), as.factor(rf_binary_actual_rush))
rf_conf_matrix_rec <- confusionMatrix(as.factor(rf_binary_predictions_rec), as.factor(rf_binary_actual_rec))

print(rf_conf_matrix_rush)
print(rf_conf_matrix_rec)

```

XGBoost MODEL
```{r}

xg_model_rush <- train(
  total_rushing_yards ~ avg_rushing_yards_per_game + opponent,
  data = train_data_rush,
  method = "xgbTree",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = expand.grid(
    nrounds = 100,          
    max_depth = 6,          
    eta = 0.3,              
    gamma = 0,              
    colsample_bytree = 0.8, 
    min_child_weight = 1,   
    subsample = 0.8         
  )
)

xg_model_rec <- train(
  total_receiving_yards ~ avg_receiving_yards_per_game + opponent,
  data = train_data_rec,
  method = "xgbTree",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = expand.grid(
    nrounds = 100,          
    max_depth = 6,          
    eta = 0.3,              
    gamma = 0,              
    colsample_bytree = 0.8, 
    min_child_weight = 1,   
    subsample = 0.8         
  )
)

```

```{r}

xg_predictions_rush <- predict(xg_model_rush, test_data_rush)

xg_mae_rush <- mean(abs(xg_predictions_rush - test_data_rush$total_rushing_yards))
print(paste("Mean Absolute Error for Rushing Yards:", xg_mae_rush))

xg_predictions_rec <- predict(rf_model_rec, test_data_rec)

xg_mae_rec <- mean(abs(xg_predictions_rec - test_data_rec$total_receiving_yards))
print(paste("Mean Absolute Error for Receiving Yards:", xg_mae_rec))

```

```{r}

rushing_threshold <- 50
receiving_threshold <- 50

xg_binary_predictions_rush <- ifelse(xg_predictions_rush > rushing_threshold, 1, 0)
xg_binary_actual_rush <- ifelse(test_data_rush$total_rushing_yards > rushing_threshold, 1, 0)

xg_binary_predictions_rec <- ifelse(xg_predictions_rec > receiving_threshold, 1, 0)
xg_binary_actual_rec <- ifelse(test_data_rec$total_receiving_yards > receiving_threshold, 1, 0)

xg_conf_matrix_rush <- confusionMatrix(as.factor(xg_binary_predictions_rush), as.factor(xg_binary_actual_rush))
xg_conf_matrix_rec <- confusionMatrix(as.factor(xg_binary_predictions_rec), as.factor(xg_binary_actual_rec))

print(xg_conf_matrix_rush)
print(xg_conf_matrix_rec)

```
DECISION TREE MODEL
```{r}
dt_model_rush <- train(
  total_rushing_yards ~ avg_rushing_yards_per_game + opponent,
  data = train_data_rush,
  method = "rpart",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = expand.grid(
    cp = seq(0.01, 0.1, by = 0.01)  
  )
)

dt_model_rec <- train(
  total_receiving_yards ~ avg_receiving_yards_per_game + opponent,
  data = train_data_rec,
  method = "rpart",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = expand.grid(
    cp = seq(0.01, 0.1, by = 0.01)
  )
)

```

```{r}
dt_predictions_rush <- predict(dt_model_rush, test_data_rush)

dt_mae_rush <- mean(abs(dt_predictions_rush - test_data_rush$total_rushing_yards))
print(paste("Mean Absolute Error for Rushing Yards:", dt_mae_rush))

dt_predictions_rec <- predict(dt_model_rec, test_data_rec)

dt_mae_rec <- mean(abs(dt_predictions_rec - test_data_rec$total_receiving_yards))
print(paste("Mean Absolute Error for Receiving Yards:", dt_mae_rec))
```

```{r}

rushing_threshold <- 50
receiving_threshold <- 50

dt_binary_predictions_rush <- ifelse(dt_predictions_rush > rushing_threshold, 1, 0)
dt_binary_actual_rush <- ifelse(test_data_rush$total_rushing_yards > rushing_threshold, 1, 0)

dt_binary_predictions_rec <- ifelse(dt_predictions_rec > receiving_threshold, 1, 0)
dt_binary_actual_rec <- ifelse(test_data_rec$total_receiving_yards > receiving_threshold, 1, 0)

dt_conf_matrix_rush <- confusionMatrix(as.factor(dt_binary_predictions_rush), as.factor(dt_binary_actual_rush))
dt_conf_matrix_rec <- confusionMatrix(as.factor(dt_binary_predictions_rec), as.factor(dt_binary_actual_rec))

print(dt_conf_matrix_rush)
print(dt_conf_matrix_rec)

```


As shown by the confusing matrices, the Random Forrest model is roughly the same 
as the XGBoost model for receiving yards, but Random Forrest is slightly better
for rushing yards, so we will use the RF Model for our predictions. But the 3rd 
model gives us our highest Accuracy and our Lowest MAE for Rushing and Receiving,
BUT it groups all the players into classes so we have predictions that are exactly the same for 2
players that are in the same class. We are going to continue to use the RF model,
because it gives us a more personalized approach for each player, without sacrificing
much accuracy.

```{r}

# Create a data frame with new player data
new_player_data <- data.frame(
  player_name = c("J.Waddle", "C.Ridley", "D.Metcalf", "A.St. Brown",
                  "D.Achane", "T.Pollard", "K.Walker", "D.Montgomery"),
  opponent = c("TEN", "MIA", "DET", "SEA",
               "TEN", "MIA", "DET", "SEA"),
  position = c("WR", "WR", "WR", "WR", "RB", "RB", "RB", "RB")
)

# Filter player averages for the new player data
player_avg_data <- player_averages %>%
  filter(player_name %in% new_player_data$player_name)

# Combine new player data with their averages
new_player_data <- merge(new_player_data, player_avg_data, by = "player_name")

# Predict rushing and receiving yards
predicted_rushing_yards <- predict(rf_model_rush, new_player_data)
predicted_receiving_yards <- predict(rf_model_rec, new_player_data)

# Print predicted yards
print(paste("Predicted Rushing Yards:", predicted_rushing_yards))
print(paste("Predicted Receiving Yards:", predicted_receiving_yards))

# Add predictions to the data frame
predictions_df <- new_player_data %>%
  mutate(
    predicted_rushing_yards = predicted_rushing_yards,
    predicted_receiving_yards = predicted_receiving_yards
  )

vegas_lines <- c(74.5, 50.5, 92.5, 65.5, 75.5, 37.5, 67.5, 71.5)

predictions_df$vegas_lines <- vegas_lines

colnames(predictions_df) <- c("Name", "Opponent", "Position", "Avg. Rushing Yards", "Avg. Receiving Yards", "Predicted Rushing Yards", "Predicted Receiving Yards", "Vegas Lines")

print(predictions_df)

table_visual <- gt(predictions_df) %>%
  tab_header(
    title = "Player Predictions",
    subtitle = "Predicted Rushing and Receiving Yards"
  ) %>%
  fmt_number(
    columns = vars(`Avg. Rushing Yards`, `Avg. Receiving Yards`, `Predicted Rushing Yards`, `Predicted Receiving Yards`, `Vegas Lines`),
    decimals = 2
  ) %>%
  cols_label(
    Name = "Player Name",
    Opponent = "Opponent",
    Position = "Position",
    `Avg. Rushing Yards` = "Avg. Rushing Yards",
    `Avg. Receiving Yards` = "Avg. Receiving Yards",
    `Predicted Rushing Yards` = "Predicted Rushing Yards",
    `Predicted Receiving Yards` = "Predicted Receiving Yards",
    `Vegas Lines` = "Vegas Lines"
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "lightblue"),
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(everything())
  ) %>%
  tab_options(
    table.font.size = 12,
    table.width = pct(100)
  )

print(table_visual)


```


player headshots + logos
```{r}
logos <- nflfastR::teams_colors_logos

aggregated_team_averages <- player_data %>% 
  group_by(opponent) %>% 
  summarize(
    rec_yds_per_game = (sum(total_receiving_yards))/21,
    rush_yds_per_game = (sum(total_rushing_yards))/21
  )

aggregated_team_averages <- merge(aggregated_team_averages, logos, by.x = "opponent", by.y = "team_abbr")

```

Rushing vs. Receiving Yards Given Up
```{r}
# g_1 <- ggplot(aggregated_team_averages, # Set dataset
#               aes(x = rec_yds_per_game, y = rush_yds_per_game)) + # Set aesthetics
#   geom_point(alpha = 0.3) +  # Set geom point
#   geom_image(image = aggregated_team_averages$team_logo_espn, asp = 16/9) + # Add logos
#   labs(y = "Average Rush Yards Given Up Per Game", # Add labels
#        x = "Average Receiving Yards Given Up Per Game",
#        title = "Receiving/Rush Yards Given Up",
#        subtitle = "NFL '23 + '24 Seasons") +
#   theme_minimal()+
#    theme(
#     panel.background = element_rect(fill = "white"),
#     plot.background = element_rect(fill = "white")
#    )
# 
# 
# g_1
```

```{r}
#ggsave("yards_given_up.png", plot = g_1, width = 10, height = 8, dpi = 300)
```


```{r}

# nfl_teams <- c("ARI", "ATL", "BAL", "BUF", "CAR", "CHI", "CIN", "CLE", "DAL", "DEN", 
#                "DET", "GB", "HOU", "IND", "JAX", "KC", "LV", "LAC", "LA", "MIA", 
#                "MIN", "NE", "NO", "NYG", "NYJ", "PHI", "PIT", "SEA", "SF", "TB", "TEN", "WAS")
# 
# player_name <- "D.Henry"
# 
# new_player_data <- data.frame(
#   player_name = rep(player_name, length(nfl_teams)),
#   opponent = nfl_teams
# )
# 
# player_avg_data <- player_averages %>%
#   filter(player_name == player_name)
# 
# new_player_data <- new_player_data %>%
#   left_join(player_avg_data, by = "player_name")
# 
# predicted_rushing_yards <- predict(rf_model_rush, new_player_data)
# predicted_receiving_yards <- predict(rf_model_rec, new_player_data)
# 
# derrick_henry_predictions <- new_player_data %>%
#   mutate(
#     predicted_rushing_yards = predicted_rushing_yards,
#     predicted_receiving_yards = predicted_receiving_yards
#   )
# 
# print(derrick_henry_predictions)

```

```{r}
# 
# derrick_henry <- merge(derrick_henry_predictions, logos, by.x = "opponent", by.y = "team_abbr")
# 
# dh_plot <- ggplot(derrick_henry, aes(x = predicted_receiving_yards, y = predicted_rushing_yards)) +
#   geom_image(aes(image = team_logo_espn), size = 0.1) +
#   labs(
#     title = "Predicted Receiving vs. Rushing Yards for Derrick Henry",
#     x = "Predicted Receiving Yards",
#     y = "Predicted Rushing Yards"
#   ) +
#   theme_minimal()+
#   theme(
#     panel.background = element_rect(fill = "white"),
#     plot.background = element_rect(fill = "white")
#   )
# print(dh_plot)
```

```{r}
#ggsave("derrick_henry.png", plot = dh_plot, width = 10, height = 8, dpi = 300)
```


```{r}

# player_name <- "S.Barkley"
# 
# new_player_data <- data.frame(
#   player_name = rep(player_name, length(nfl_teams)),
#   opponent = nfl_teams
# )
# 
# player_avg_data <- player_averages %>%
#   filter(player_name == player_name)
# 
# new_player_data <- new_player_data %>%
#   left_join(player_avg_data, by = "player_name")
# 
# predicted_rushing_yards <- predict(rf_model_rush, new_player_data)
# predicted_receiving_yards <- predict(rf_model_rec, new_player_data)
# 
# saquon_predictions <- new_player_data %>%
#   mutate(
#     predicted_rushing_yards = predicted_rushing_yards,
#     predicted_receiving_yards = predicted_receiving_yards
#   )
# 
# print(saquon_predictions)

```
```{r}
# saquon_barkley <- merge(saquon_predictions, logos, by.x = "opponent", by.y = "team_abbr")
# 
# sb_plot <- ggplot(saquon_barkley, aes(x = predicted_receiving_yards, y = predicted_rushing_yards)) +
#   geom_image(aes(image = team_logo_espn), size = 0.10) +
#   labs(
#     title = "Predicted Receiving vs. Rushing Yards for Saquon Barkley",
#     x = "Predicted Receiving Yards",
#     y = "Predicted Rushing Yards"
#   ) +
#   theme_minimal()+
#     theme(
#     panel.background = element_rect(fill = "white"),
#     plot.background = element_rect(fill = "white")
#   )
# print(sb_plot)
```

```{r}
#ggsave("saquon_barkley.png", plot = sb_plot, width = 10, height = 8, dpi = 300)
```


```{r}

# player_name <- "J.Jefferson"
# 
# new_player_data <- data.frame(
#   player_name = rep(player_name, length(nfl_teams)),
#   opponent = nfl_teams
# )
# 
# player_avg_data <- player_averages %>%
#   filter(player_name == player_name)
# 
# new_player_data <- new_player_data %>%
#   left_join(player_avg_data, by = "player_name")
# 
# predicted_rushing_yards <- predict(rf_model_rush, new_player_data)
# predicted_receiving_yards <- predict(rf_model_rec, new_player_data)
# 
# jettas_predictions <- new_player_data %>%
#   mutate(
#     predicted_rushing_yards = predicted_rushing_yards,
#     predicted_receiving_yards = predicted_receiving_yards
#   )
# 
# print(jettas_predictions)

```
```{r}
# justin_jefferson <- merge(jettas_predictions, logos, by.x = "opponent", by.y = "team_abbr")
# 
# jj_plot <- ggplot(justin_jefferson, aes(x = predicted_receiving_yards, y = predicted_rushing_yards)) +
#   geom_image(aes(image = team_logo_espn), size = 0.1) +
#   labs(
#     title = "Predicted Receiving vs. Rushing Yards for Justin Jefferson",
#     x = "Predicted Receiving Yards",
#     y = "Predicted Rushing Yards"
#   ) +
#   theme_minimal()+
#     theme(
#     panel.background = element_rect(fill = "white"),
#     plot.background = element_rect(fill = "white")
#   )
# print(jj_plot)
```

```{r}
#ggsave("justin_jefferson.png", plot = jj_plot, width = 10, height = 8, dpi = 300)
```


```{r}

# player_name <- "J.Chase"
# 
# new_player_data <- data.frame(
#   player_name = rep(player_name, length(nfl_teams)),
#   opponent = nfl_teams
# )
# 
# player_avg_data <- player_averages %>%
#   filter(player_name == player_name)
# 
# new_player_data <- new_player_data %>%
#   left_join(player_avg_data, by = "player_name")
# 
# predicted_rushing_yards <- predict(rf_model_rush, new_player_data)
# predicted_receiving_yards <- predict(rf_model_rec, new_player_data)
# 
# jamaar_chase_predictions <- new_player_data %>%
#   mutate(
#     predicted_rushing_yards = predicted_rushing_yards,
#     predicted_receiving_yards = predicted_receiving_yards
#   )
# 
# print(jamaar_chase_predictions)

```

```{r}
# jamaar_chase <- merge(jamaar_chase_predictions, logos, by.x = "opponent", by.y = "team_abbr")
# 
# jc_plot <- ggplot(jamaar_chase, aes(x = predicted_receiving_yards, y = predicted_rushing_yards)) +
#   geom_image(aes(image = team_logo_espn), size = 0.1) +
#   labs(
#     title = "Predicted Receiving vs. Rushing Yards for Ja'Marr Chase",
#     x = "Predicted Receiving Yards",
#     y = "Predicted Rushing Yards"
#   ) +
#   theme_minimal()+
#     theme(
#     panel.background = element_rect(fill = "white"),
#     plot.background = element_rect(fill = "white")
#   )
# print(jc_plot)
```

```{r}
#ggsave("jamaar_chase.png", plot = jc_plot, width = 10, height = 8, dpi = 300)
```


SINGLE PLAYER PREDICTION
```{r}
# Input data for one player and the opponent he is facing
single_player_data <- data.frame(
  player_name = "J.Dobbins"  ,
  opponent = "JAX"
)

# Filter player averages for the single player data
player_avg_data <- player_averages %>%
  filter(player_name == single_player_data$player_name)

# Combine single player data with their averages
single_player_data <- merge(single_player_data, player_avg_data, by = "player_name")

# Predict rushing and receiving yards
predicted_rushing_yards <- predict(rf_model_rush, single_player_data)
predicted_receiving_yards <- predict(rf_model_rec, single_player_data)

# Print predicted yards
print(paste("Predicted Rushing Yards:", predicted_rushing_yards))
print(paste("Predicted Receiving Yards:", predicted_receiving_yards))

```

