---
title: "Predicting UFC Fights"
date: "07/31/2024"
author: Paul Trusela
output: 
  slidy_presentation: default
  beamer_presentation: default
  iosldies_presentation: default
---
```{r boring, include=FALSE}
library(dplyr)
library(sjPlot)
library(ggplot2)
library(predict3d)
library(corrr)
library(DALEX)
library(lme4)
library(car)
```


```{r Cleaning Data, echo=TRUE}

ufc_fight_data <- read.csv("C:/users/pauly/downloads/fight_data.csv")
ufc_fighter_data <- read.csv("C:/users/pauly/downloads/official_fighter_data.csv")


ufc_fight_data$win <- ifelse(ufc_fight_data$res == "W", 1, 0)

ufc_agg <- ufc_fight_data %>% 
  group_by(full_name) %>% 
  summarise(wins = sum(win),
            fights = n(),
            AKDL = mean(kd),
            AKDR = mean(kds_received),
            TSL = mean(strike_landed),
            ASRR = mean(sig_reg_percent)
            )


ufc_agg <- left_join(ufc_agg,ufc_fighter_data, by="full_name")

ufc_agg$win_rate <- ufc_agg$wins/ufc_agg$fights

ufc_agg <- na.omit(ufc_agg)

```
Can we use a select few UFC stats to try to predict win rates, and
winners of fights?

I scraped my data from ufcstats.com and had all sorts of stats, over 30+, I mutated
and transformed my data into something I could work with and then started checking the correlation
of certain variables with others.

```{r Correlation Plot, echo=TRUE, message=FALSE, warning=FALSE}

ufc_correlate <- correlate(ufc_agg)

rplot(ufc_correlate) +
  theme(axis.text.x = element_text(angle = 90))

```


AKDL - Average Knockdowns Landed per Fight

AKDR - Average Knockdowns Received per Fight

TSL - Total Strikes Landed

ASRR - Average Significant to Regular strike Ratio

SLPM - Significant Strikes Landed per Minute

SAPM - Significant Strikes Absorbed per Minute

STRD - Significant Strike Defense (the % of opponents strikes that did not land)

TD - Average Takedowns Landed per 15 minutes

TDA - Takedown Accuracy (% of TD landed)

TDD - Takedown Defense (the % of opponents TD attempts that did not land)

SUBA - Submission Attempts




After Checking the correlation of the variables, I started playing around with the
model to try and figure out which stats would be most important to use, I ultimatley
settled on 9 different stats that cover variopus aspects of a UFC fight.

```{r Model, echo=TRUE, warning=FALSE}
ufc_model <- lm(win_rate ~ 
                  AKDL +
                  AKDR +
                  SLPM +
                  SAPM +
                  STRD +
                  TDD +
                  ASRR +
                  TD +
                  TSL
                  ,data = ufc_agg)

summary(ufc_model)
```
Overall I was moderately happy with how the model turned out, the P-value is very
low, and the R-squared is decent, definitely could be better. But I would say there 
is a moderate correlation among the variables chosen and the model could be used
to get a baseline idea of who might win a fight, and the model could also even show
fighters what areas are most important to work on, as seen later in the file.
All the variables are statistically significant, with AKDR + SAPM being the only negative 
correlations.

```{r}
performance::check_collinearity(ufc_model)
```
Checking the VIF scores, they all seem to be relatively low, the only one that
may raise a small amount of concern is the SLPM or Significant Strikes Landed per Minute, 
but I believe that to be a very important stat to include in our model.

```{r echo=TRUE, message=FALSE, warning=FALSE}
set.seed(1001)

train_indices <- sample(1:nrow(ufc_agg), 
                        0.8 * nrow(ufc_agg), 
                        replace = FALSE)

train_data <- ufc_agg[train_indices, 
                c("win_rate",
                  "AKDL",
                  "AKDR",
                  "SLPM", 
                  "SAPM",
                  "STRD",
                  "TDD", 
                  "ASRR",
                  "TD",
                  "TSL")]

train_data <- na.omit(train_data)

test_data <- ufc_agg[-train_indices, 
                c("win_rate",
                  "AKDL",
                  "AKDR",
                  "SLPM", 
                  "SAPM",
                  "STRD",
                  "TDD", 
                  "ASRR",
                  "TD",
                  "TSL")]

test_data <- na.omit(test_data)

train_ufc_model <- lm(win_rate ~ 
                  AKDL +
                  AKDR +
                  SLPM +
                  SAPM +
                  STRD +
                  TDD +
                  ASRR +
                  TD +
                  TSL
                  ,data = ufc_agg)

test_data$predicted_win_rate <- predict(train_ufc_model, 
                                   newdata = test_data)


```
Next I am developing train and test data to use for the prediction portion of the
model, which will ultimately be used to predict an outcome of a fight between 2 fighters.

```{r echo=TRUE}

train_explain <- explain(
  train_ufc_model, 
  data = train_data[, which(names(train_data) != "win_rate")], 
  y = train_data$win_rate,
)

```

```{r train performance, echo=TRUE}
train_performance <- model_performance(train_explain)

train_performance
```
Although the R-Squared could definitley be improved, the model overall is still
statistically significant.

```{r Predicted Test Data, echo=TRUE}
shap_plot <- predict_parts(
  train_explain, 
  new_observation = test_data, 
  type = "shap")

plot(shap_plot)
```

The plot above shows how a fighter's win percentage would be effected based on various 
values for each variable.


```{r}
model_var_imp <- model_parts(train_explain, type = "variable_importance")

plot(model_var_imp)
```
We see that the least important variable in the model is the STRD or Significant 
Strike Defense, which I was surprised at since its pretty important to try to do
dodge the punches being thrown at you.



```{r Variable effects on Win Percentage, echo=TRUE, message=FALSE, warning=FALSE}
model_var_profile <- variable_profile(train_explain, type = "conditional")

plot(model_var_profile, title = " ")
```

This graph is a good visual to show how drastically a variable effects
a fighter's win percentage. with Average Knockdowns Landed (AKDL), 
being most drastic.

```{r Break Down of Model, echo=TRUE}
break_down_plot <- predict_parts(
  train_explain, 
  new_observation = test_data, 
  type = "break_down")

plot(break_down_plot, title = "Break Down Profile of Model")
```

```{r Jon Jones vs Stipe Miocic, echo=TRUE, message=FALSE, warning=FALSE}
jon_jones <- tibble(
                  AKDL = 0.23,
                  AKDR = 0,
                  SLPM = 4.29,
                  SAPM = 2.22,
                  STRD = 0.64,
                  TDD = 0.95,
                  ASRR = 0.62,
                  TD = 1.93,
                  TSL = 83.41
)

predict(
    train_ufc_model,
    newdata = jon_jones
)

stipe_miocic <- tibble(
                  AKDL = 0.39,
                  AKDR = 0.28,
                  SLPM = 4.82,
                  SAPM = 3.82,
                  STRD = 0.54,
                  TDD = 1,
                  ASRR = 0.59 ,
                  TD = 1.86,
                  TSL = 85.39
)

predict(
    train_ufc_model,
    newdata = stipe_miocic
)
```

Now I am testing the projected win rate for 2 fighters based on their stats
inputed into the model. According to the model Jon Jones has a slightly higher
expected win rate, but this does not take into account the opponent that each would
be fighting.

```{r Jon Jones Break Down, echo=TRUE}
break_down_plot_jon_jones <- predict_parts(
  train_explain, 
  new_observation = jon_jones, 
  type = "break_down")

plot(break_down_plot_jon_jones, title = "Break Down Profile for Jon Jones")
```

Above is a breakdown of Jon Jones stats and hiow they affect his projected win rate.

```{r Stipe Miocic Break Down, echo=TRUE}
break_down_plot_stipe_miocic <- predict_parts(
  train_explain, 
  new_observation = stipe_miocic, 
  type = "break_down")

plot(break_down_plot_stipe_miocic, title = "Break Down Profile for Stipe Miocic")
```

Above is a visual for Stipe Miocic's stats, and how they affect his win rate.
The red being what negatively affects him, strikes absorbed per minute and knockdowns
received, could provide insights into areas a fighter may need to improve.

Overall I am happy with how the model turned out. I don't necessarily trust it as
a predictor to bet money with, but I think the visuals that came out of it can be
super helpful for coaches and for fans to see. I think I could improve the model by
somehow integrating fighters age, and/or the level of opponents they have face, 
since those would likely have a large impact on the prediction side of the model.
Another improvement could include developing a way to compare 2 fighters directly 
to each other as a better way of predicting a fight based on how their stats 
match-up.
